EagerNet: Early Predictions in Feedforward Neural Networks for Fast Network Intrusion Detection

### 2020-07-20 paper/paper.tex
add content to paper

\section{Introduction}
% what are ffnn
\glspl{ffnn} has gained remarkable attention in recent years due to the availability of data and computational resources for research. Many of the state-of-the-art architectures currently used in different applications are effectively based on simple \glspl{ffnn} architectures. Experiments show that the more layers (and neurons) the network includes, the more desirable it is for better learning \cite{eldan2016power} .

@@

@@

@@

@@

	\centering
	\caption{CAIA network flow features.}
	\label{tab:cres}

	\footnotesize
	\begin{tabular}{|c|c|c|}
		\toprule
		\textbf{Direction}                    & \textbf{Features}        & \textbf{Statistical Operations}        \\
		\midrule

@@

@@

\begin{figure*}[htp]
\center
\subfloat[Conventional \gls{ffnn} architecture.]{%
  \includegraphics[clip,width=0.45\textwidth]{figures/ffnn_new.pdf}%
  \label{fig:ffnn}
}
\\
\subfloat[EagerNet: Forward-pass.]{%
  \includegraphics[clip,width=0.53\textwidth]{figures/eagernet_new.pdf}%
}
\quad
\subfloat[EagerNet: Back-propagation of gradients.]{%
  \includegraphics[clip,width=0.45\textwidth]{figures/eagernet_backprop_new.pdf}%
}

@@

\subsubsection{Binary Vs. Multiclass Classification}

\begin{equation}
l = \frac{1}{N} \sum_{n}^{N} -\omega [y_n \cdot \log \sigma (x_{n}) + (1 - y_n) \cdot \log (1 - \sigma (x_{n}))]
\end{equation}

\[
\nabla l = \underbrace{W_{current} \times \nabla l_{current}}_{\text{Current gradient}} + \underbrace{\sum_{i}^{} \nabla l_{i}}_{\text{Next layer gradients}}
\]

\subsubsection{Combined Back-propagation Loss}

\subsubsection{Design Decision}

@@

\section{Evaluation and Discussion}
\subsection{Datasets}

@@

\begin{table*}
\parbox{.45\linewidth}{
\centering
\begin{tabular}{cccrrrr}
\toprule
\textbf{Variant} & \textbf{Weights} & \textbf{Layers} & \textbf{Acc.} & \textbf{Prec.} & \textbf{Rec.} & \textbf{J.}\\
\midrule
\multirow{6}{*}{\rotatebox{90}{Binary}} & \multirow{2}{*}{Equal} & 1-3-1 & 0.996 & 0.996 & 0.990 & 0.989 \\
 & & 1-1-1 & 0.997 & 0.997 & 0.990 & 0.990 \\
 & \multirow{2}{*}{Increasing} & 1-3-1 & 0.997 & 0.998 & 0.990 & 0.989 \\
 & & 1-1-1 & 0.997 & 0.997 & 0.990 & 0.989 \\
 & \multirow{2}{*}{Decreasing} & 1-3-1 & 0.997 & 0.998 & 0.989 & 0.989 \\
 & & 1-1-1 & 0.997 & 0.997 & 0.990 & 0.990 \\
\midrule
\multirow{6}{*}{\rotatebox{90}{Multiclass}} & \multirow{2}{*}{Equal} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Increasing} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Decreasing} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
\end{tabular}
\caption{CIC-IDS17}
}
\hfill
\parbox{.45\linewidth}{
\centering
\begin{tabular}{cccrrrr}
\toprule
\textbf{Variant} & \textbf{Weights} & \textbf{Layers} & \textbf{Acc.} & \textbf{Prec.} & \textbf{Rec.} & \textbf{J.}\\
\midrule
\multirow{6}{*}{\rotatebox{90}{Binary}} & \multirow{2}{*}{Equal} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Increasing} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Decreasing} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
\midrule
\multirow{6}{*}{\rotatebox{90}{Multiclass}} & \multirow{2}{*}{Equal} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Increasing} & 1-3-1 & & & & \\
 & & 1-1-1 & & & & \\
 & \multirow{2}{*}{Decreasing} & 1-3-1 & 0.988 & 0.846 & 0.830 & 0.825 \\
 & & 1-1-1 & & & & \\
\end{tabular}
\caption{UNSW-NB15}
}
\end{table*}

@@

\centering
\begin{tabular}{ccrr}
\toprule
\textbf{Dataset} & \textbf{Architecture} & \textbf{FFNN} & \textbf{EagerNet*}\\
\midrule
\multirow{6}{*}{\rotatebox{90}{CIC-IDS17}} & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
\midrule
\multirow{6}{*}{\rotatebox{90}{UNSW-NB15}} & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
\midrule
\multirow{6}{*}{\rotatebox{90}{MNIST}} & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
 & & & \\
\end{tabular}
\vspace{1ex}

{\raggedright * . \par}
\caption{Accuracy scores.}

@@

\subsection{Confidence-Speed Tradeoff}

\begin{figure}[b!]
\includegraphics[width=\columnwidth]{figures/conf-acc-layers.pdf}
\caption{Accuracy-Confidence plot.\note{red}{FM: this will change}}
\label{fig:pdp_ttl}
\end{figure}

\begin{figure}[b!]
\includegraphics[width=\columnwidth]{figures/acc_multiclass_Jul10_15-18-24_gpu_0_3.pdf}
\caption{Accuracy plot multiclass.\note{red}{FM: this will change}}
\label{fig:pdp_ttl}
\end{figure}

>>>>

\section{Introduction}
% what are ffnn
\glspl{ffnn} have gained remarkable attention in recent years due to the availability of data and computational resources for research. Many of the state-of-the-art architectures currently used in different applications are effectively based on simple \glspl{ffnn} architectures. Experiments show that the more layers (and neurons) the network includes, the more desirable it is for better learning \cite{eldan2016power} .

@@

% talk about network traffic

@@

 \begin{figure}
  \includegraphics[width=\linewidth]{figures/ffnn_new.pdf}
  \caption{Conventional \gls{ffnn} architecture.}
  \label{fig:ffnn}
\end{figure}

@@

\gls{ml} has been widely used in the last decade for network traffic analysis and specifically anomaly detection systems. Many of the works proposed \gls{ids} architectures based on well-known supervised techniques \cite{survey}. The idea of training a classifier on pre-stored attack patterns and using the same classifier to recognize similar patterns is commonly used and remarkable success has so far been achieved \cite{survey}. In particular, \glspl{ffnn} is commonly used whenever large amounts of data are available. In order to present network data to such architectures, a number of traffic representation families have been proposed depending on the application. However, the statistical representation of flow characteristics is still the most appropriate. In this work, we define the network flow to be the exchange of packets between two end-hosts. Packets can be identified and aggregated using the five-tuple key: \emph{sourceIP}, \emph{destinationIP}, \emph{sourceTransportPort}, \emph{destinationTransportPort} and \emph{protocolIdentifier}. Thereafter, packet features are extracted and the statistical combinations are computed. Table \ref{tab:features} shows CAIA \cite{williams2006preliminary} network traffic representation. It consists of 12 features 7 of which are measured in both directions and 2 of which are expanded using four statistical combinations (i.e. \emph{A} becomes \emph{mean(A)}, \emph{min(A)}, \emph{max(A)} and \emph{stdev(A)}). In addition, we set an observation timeout of 1,800 seconds, after which we terminate flows. Each flow is considered to be a data sample and has two sets of labels: binary (attack/benign) label and attack family. After a training process, the trained \gls{ids} must be able to correctly classify new flows.

@@

	\centering
	\caption{CAIA flow representation.}
	\label{tab:features}

	\footnotesize
	\resizebox{\columnwidth}{!}{
	\begin{tabular}{|c|c|c|}
		\toprule
		\textbf{Direction}                    & \textbf{Features}        & \textbf{Statistical Operations}        \\
		\midrule

@@

	\end{tabular}
	}

@@

Because several network attack families are simple to detect on the basis of linear input combinations, the standard \gls{ffnn} is slightly modified to cope with the difficulty of detection. Our novel architecture is based on the assumption that deep neural networks can learn complex functions with simpler mappings learned in the first few layers. As a result, we build a network such that an additional set of neurons is connected to each layer (a copy of the output layer), allowing for predictions at each layer. The EagerNet architecture is shown in Figure \ref{fig:eagerNet_forward}. Initial neurons are shown in gray and output neurons (per layer) are shown in green. Once a flow is observed and features are extracted, a sample $\textbf{X}$ is fed to the network via the input layer. The network proceeds evaluating the output of each layer and yields a prediction confidence value. We define the confidence as the value that represents how sure the network is about a certain decision or sample belonging to a certain class. Based on a predefined threshold, the network measures confidence and determines whether to avoid testing additional layers or proceed to do so in order to gain more confidence. This methodology ensures that simpler samples are classified early in the forward-pass and that more complicated, noisy samples are passed into deeper layers. In the \ref{confidence-speed-tradeoff} Section, we show that EagerNet reduces resource consumption on average keeping comparable performance to normal \gls{ffnn}.

\begin{figure*}[htp]
\center

\subfloat[EagerNet: Forward-pass.]{%
  \includegraphics[clip,width=0.5\textwidth]{figures/eagernet_new.pdf}%
  \label{fig:eagerNet_forward}
}
\quad
\subfloat[EagerNet: Back-propagation of gradients.]{%
  \includegraphics[clip,width=0.42\textwidth]{figures/eagernet_backprop_new.pdf}%
  \label{fig:eagerNet_backward}
}

@@

\subsubsection{Binary Vs. Multiclass Classification}
We adjust the output layer to build two separate sub-architectures. The first architecture, consisting of a binary attack detection, i.e. a 1 is an attack flow and a 0 is a benign flow. The predictions in this case are given by a single output neuron followed by a sigmoid activation function to force values between 0 and 1. During back-propagation, we use a binary cross-entropy loss including the activation function and average the loss over the batch. Equation \ref{eq:bceloss} shows how the loss is determined for a batch of size $N$, with the $i^{th}$ input $x_{i}$ and the sigmoid activation function  $\sigma()$.

\begin{equation}
\label{eq:bceloss}
l = \frac{1}{N} \sum_{n}^{N} -\omega [y_n \cdot \log \sigma (x_{n}) + (1 - y_n) \cdot \log (1 - \sigma (x_{n}))]
\end{equation}

In the case of a multicalss class problem, the output layers consist of as many neurons as attack families followed by a softmax activation functions that allow at most only one neuron to be 1. During back-propagation, we use the categorical cross-entropy loss to penalize all outputs together, so there is no overlap between classes. Equation \ref{eq:cceloss} shows ...

\begin{equation}
\label{eq:bceloss}
l = \frac{1}{N} \sum_{n}^{N} -\omega [y_n \cdot \log \sigma (x_{n}) + (1 - y_n) \cdot \log (1 - \sigma (x_{n}))]
\end{equation}

After computing the losses, we use $Adam$ optimizer for computing gradients with a learning rate of 0.001.

\subsubsection{Combined Back-propagation Loss}
Conventional \gls{ffnn} uses the loss of the last layer to adjust the gradients. In this research, and since we have multiple outputs; one at each layer, we aggregate the losses and back propagate them. Fig \ref{fig:eagerNet_backward} demonstrates how losses propagate back from the respective layer only. As a consequence, the weights of the last layer are affected only by the loss of the last output. Similarly, the weights of the first layer are affected by the losses of the entire network. In addition, we introduce a weighting policy for losses. We allocate a weight to each loss that is predefined prior to training. The aim is to help the network determine on which layer to optimize more. Three types of weight sets are defined:
\begin{itemize}
\item Uniform weights. all losses have the same weight:
\begin{equation}
\mathcal{W_U} = \lbrace W_{0}, ..., W_{n} \rbrace = \lbrace \frac{1}{\sum_{i=0}^{n} 1}, ..., \frac{1}{\sum_{i=0}^{n} 1} \rbrace
\end{equation}

\item Increasing weights. losses are increasingly important from first to last layer. 
\begin{equation}
\mathcal{W_I} = \lbrace W_{0}, ..., W_{n} \rbrace = \lbrace \frac{1}{\sum_{i=0}^{n} i+1}, ..., \frac{n+1}{\sum_{i=0}^{n} i+1} \rbrace
\end{equation}

\item Decreasing weights. losses are decreasingly important from first to last layer. 
\begin{equation}
\mathcal{W_D} = \lbrace W_{0}, ..., W_{n} \rbrace = \lbrace \frac{n+1}{\sum_{i=0}^{n} i+1}, ..., \frac{1}{\sum_{i=0}^{n} i+1} \rbrace
\end{equation}

\end{itemize}

We show the effect of the weight distributions in Table \ref{tab:weights} and discuss results in Section \ref{evaluation_and_discussion}.

%\begin{equation}
%\nabla l = {\underbrace{W_{current} \times \nabla l_{current}}_{\text{Current gradient}}} + {\underbrace{\sum_{i}^{} \nabla l_{i}}_{\text{Next layer gradients}}}
%\end{equation}

%\subsubsection{Enhanced Training}
%Back-propagating all losses together even when using weighted-loss summation causes the network to minimize the loss on each layer hence learning all classes at each layer simultaneously if the training procedure is long enough thus causing (1) the first few layers to over-fit and (2) the network's loss to converge slowly if there are many classes to learn. To solve this issue, we propose a different weights distribution that constructs the weights as follows:
%\begin{itemize}
%\item perform a forward-pass and compute the loss per layer
%\item select the layer with the minimal loss
%\item attributes a decreasing weight distribution starting from the selected layer until the last layer
%\end{itemize}
%The above procedure insures that the most suitable layer learns the respective pattern as well as all subsequent layers hence ignoring first layers which as discussed before, may over-fit. This allows the network to focus on one layer and class at a time and let subsequent layers to learn the same patterns in case the confidence threshold is set too high. Intuitively, easy patterns are learned in the first layers since the loss is expected to be minimal and more complex patterns are learned in the last layers because more non-linearities can be obtained.
%
%\begin{equation}
%\begin{split}
%\mathcal{W_M} & = \lbrace W_{0}, ...,W_{m-1}, \underbrace{W_{m}}_{\text{Minimum Loss Index}}, ..., W_{n} \rbrace \\
% & = \lbrace 0, ..., 0, \frac{2^{n}}{\sum_{i=m}^{n} 2^i}, ..., \frac{2^m}{\sum_{i=m}^{n} 2^i} \rbrace
% \end{split}
%\end{equation}

% easy pattern less loss, first layers
% summing up loss causes to learn the same pattern at all layers
% assume that even in binary classifications, there are many subclasses

@@

\section{Evaluation and Discussion}
\label{evaluation_and_discussion}
We discuss in this section different experimental results and observations. Overall, EagerNet shows promising performance comparable to \glspl{ffnn} with the ability to save resources.

\subsection{Datasets}

@@

% why from 512 to 32

@@

\centering

\begin{tabular}{cccrrrr}
\toprule
\textbf{Variant} & \textbf{Weights} & \textbf{Layers} & \textbf{Acc.} & \textbf{Prec.} & \textbf{Rec.} & \textbf{J.}\\
\midrule
\multirow{6}{*}{\rotatebox{90}{CIC-IDS17}} & \multirow{2}{*}{Equal} & 5 & 0.996 & 0.996 & 0.990 & 0.989 \\
 & & 3 & 0.997 & 0.997 & 0.990 & 0.990 \\
 & \multirow{2}{*}{Increasing} & 5 & 0.997 & 0.998 & 0.990 & 0.989 \\
 & & 3 & 0.997 & 0.997 & 0.990 & 0.989 \\
 & \multirow{2}{*}{Decreasing} & 5 & 0.997 & 0.998 & 0.989 & 0.989 \\
 & & 3 & 0.997 & 0.997 & 0.990 & 0.990 \\
\midrule
\multirow{6}{*}{\rotatebox{90}{UNSW-NB15}} & \multirow{2}{*}{Equal} & 5 & 0.988 & 0.828 & 0.867 & 0.860 \\
 & & 3 & 0.988 & 0.856 & 0.800 & 0.795 \\
 & \multirow{2}{*}{Increasing} & 5 & 0.988 & 0.853 & 0.821 & 0.816 \\
 & & 3 & 0.988 & 0.844 & 0.817 & 0.811 \\
 & \multirow{2}{*}{Decreasing} & 5 & 0.988 & 0.843 & 0.824 & 0.819 \\
 & & 3 & 0.988 & 0.851 & 0.819 & 0.813 \\

\vspace{1ex}

{\raggedright * All architectures consist of 128 neurons per layer in addition to single output layer. \par}
\caption{Effect of weights on the performance of different networks}
\label{tab:weights}

@@

\centering
\begin{tabular}{ccrc}
\toprule
\textbf{Dataset} & \textbf{Variant*} & \textbf{\gls{ffnn}} & \textbf{EagerNet (last)} \\
\midrule
\multirow{2}{*}{CIC-IDS17} & Binary & 0.983 & 0.988 \\
 & Multiclass & & \\
\midrule
\multirow{2}{*}{UNSW-NB15} & Binary & & \\
 & Multiclass & & \\
\midrule
MNIST & Multiclass & & \\
\midrule
\end{tabular}
\vspace{1ex}

{\raggedright * All architectures consist of 10 layers $\times$ 64 neurons in addition to input and output layers. \par}
\caption{Accuracy scores.}

\subsection{Confidence-Speed Tradeoff}
\label{confidence-speed-tradeoff}

\begin{figure*}
\center
\begin{tabular}{cccc}
\subfloat[CIC-IDS17 (3 layers $\times$ 128 neurons).]{\includegraphics[width = 2.5in]{figures/conf-acc-17-3x128.pdf}} &
\subfloat[CIC-IDS17 (12 layers $\times$ 64 neurons).]{\includegraphics[width = 2.5in]{figures/conf-acc-17-12x64.pdf}}\\
\subfloat[UNSW-NB15 (3 layers $\times$ 128 neurons).]{\includegraphics[width = 2.5in]{figures/conf-acc-15-3x128.pdf}} &
\subfloat[UNSW-NB15 (12 layers $\times$ 64 neurons).]{\includegraphics[width = 2.5in]{figures/conf-acc-15-12x64.pdf}}
\end{tabular}
\caption{Confidence threshold effect on accuracy and number of needed layers.}
\label{fig:confidence}
\end{figure*}

\begin{figure}
\center
\begin{tabular}{cccc}
\subfloat[CIC-IDS17 (12 layers $\times$ 64 neurons).]{\includegraphics[width = 3in]{figures/multiclass17.pdf}} &
\subfloat[UNSW-NB15 (12 layers $\times$ 64 neurons).]{\includegraphics[width = 3in]{figures/multiclass17.pdf}}
\end{tabular}
\caption{Accuracy plot multiclass.}
\end{figure}

\subsubsection{Optimal Confidence Threshold}
During deployment, the optimum confidence threshold is set by the user and plays a role in the overall achieved accuracy. Setting the threshold too low can cause the network to make decisions at an early stage, thereby gaining resources but achieving low accuracy. Similarly, setting the threshold too high can cause the network to skip the proper layers and thus always use the last one. The desired confidence level can therefore be obtained during the training phase by determining on the basis of the desired accuracy as shown in Figure \ref{fig:confidence}. % why
